---
title: "Part 3: Generalized Linear Models"
author: "Charles Julien, Chike Odenigbo, Atul Sharma, Gabriel Jobert"
date: "11/17/2023"
geometry: "left=2cm,right=2cm,top=1cm,bottom=1.2cm"
output:
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, warning=FALSE, include = FALSE, echo=FALSE}
# Import (The fewer the better)
#install.packages('AER')
library(tidyverse)
library(ggplot2)
library(dplyr)
library(leaflet)
library(gridExtra)
library(broom)
library(stats)
library(car)
library(janitor)
library(tibble)
library(MASS)
library(knitr)
library(AER)
```

```{r load_data, include=FALSE}
# Data loading

# Set working directory to file location

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
df_raw = read.csv("./../src/bixi10.csv")
df_station = read.csv("./../src/2021_stations.csv")
df_main = left_join(df_raw,df_station,by = c("station"="pk"))
head(df_main)

```

```{r echo=FALSE}
# Data preparation and feature engineering


# Note: Correlation between rev and dur is very high 0.994467

df_main <- df_main %>% mutate(
              holiday = factor(holiday),
              mem = factor(mem),
              mm = factor(mm),
              station = factor(station),
              
              wday_ordered = factor(wday, levels = c("Monday", "Tuesday", "Wednesday",  "Thursday", "Friday", "Saturday", "Sunday"),ordered = TRUE),
              season = ifelse((mm %in% c(12,1,2)), 'Winter',
                       ifelse((mm %in% c(3,4,5)), 'Spring',
                       ifelse((mm %in% c(6,7,8)), 'Summer',
                       ifelse((mm %in% c(9,10,11)), 'Fall','No Season')))),
              
              rain_ind = factor(ifelse((rain != 0), 'Rain', 'NoRain')),
              
              # Compared to Parc Lafontaine 
              North_South = factor(ifelse(latitude>45.5271, 'North', 'South')),
              West_East = factor(ifelse(longitude>-73.5705, 'East', 'West')),
              
              # Beginning of month versus end of month
              PartOfMonth = factor(ifelse(dd>15, 'EOM', 'BOM')),
              
              Metro_ind = factor(ifelse(grepl("Métro", name),1,0)),
              
              lntot = log(n_tot),
              
              rev_per_min = rev/dur,
              rev_per_trip = rev/n_tot,
             
              percent_AM = n_AM/n_tot,
              percent_PM = n_PM/n_tot,
              
              percent_AM_PM_delta = percent_AM - percent_PM,
              n_AM_PM_delta = n_AM - n_PM,
              
              am_pm_ind = ifelse(n_AM>=n_PM, 'AM','PM'),
              
              wknd_ind = factor(ifelse((wday %in% c("Sunday","Saturday")), 'Weekend'
                        ,'Weekday')),
                
              long_wknd_ind = factor(ifelse((wday %in% c("Sunday","Saturday")), 'Weekend'
                        ,ifelse((wday %in% c("Friday","Monday") & holiday=='1'),'Long Weekend','Weekday'))),
        
              year = 2021 ,
              
              date = as.Date(paste(year,mm,dd,sep="/")),
              
              week_num = as.numeric(strftime(date, format = "%V")),
 
              hot_weather_ind = factor(ifelse(temp > 15, 'Hot', 'Cold')),
# create a new variable indicating whether the average daily trip duration exceeds 15 minutes, and explore models for this new variable.

              avg_15_ind = factor(ifelse(avg>15,1,0))

)


```

# Introduction

In the dynamic realm of urban mobility, the Bixi public cycling service
plays a pivotal role in providing a sustainable and accessible
transportation alternative. As consultants entrusted with a
comprehensive analysis of Bixi's operational data, our approach
integrates sophisticated statistical techniques, specifically
Generalized Linear Models (GLM), to derive actionable insights. This
report unfolds the findings derived from GLM applications, shedding
light on critical aspects such as trip durations, ridership patterns,
and the impact of external factors.

Our analytical scope encompasses a multifaceted examination of factors
influencing trip durations, total trip counts, and the temporal dynamics
of ridership. By applying GLM to address the identified research
questions, we aim to unearth insights that are instrumental in shaping
strategic decisions for Bixi's operational enhancements. Central to our
methodology is the implementation of Generalized Linear Models, a
statistical framework adept at capturing complex relationships within
diverse datasets. Our application of GLM is tailored to respond to
specific research questions, providing a granular understanding of the
nuanced dynamics at play in Bixi's operational landscape.

The main focus of this analysis will be on number of rentals (total, AM,
and PM), and on long trips (\>15min).

## Preliminary T-Test for Variable Selection

As a general hypothesis, prior to diving into our research questions we
explored the target variables through variables of interest using
t-tests. More specifically, we wanted to determine whether the holiday
period and the weather had a significant impact on the targets.

Using a 2 sample t-test, we tested if the average trip duration was the
same in holiday vs non holiday period. In all cases, the test statistic
is the mean difference difference by the difference in standard
deviations of both groups. Using a significance level of 5% we can
reject the null hypothesis and conclude that the average trip duration
is not the same in holiday and non-holiday periods though they differ by
less than a minute using the sample means of the groups.

```{r}
t.test(avg ~ holiday, data = df_main, var.equal = TRUE)
```

On the other hand, looking at the total number of trips we can conclude
that using a significance level of 5% that the trips do not differ from
holiday periods to non-holiday periods.

```{r}
t.test(n_tot ~ holiday, data = df_main, var.equal = TRUE)
```

Looking at membership, we can conclude that non-members take longer
trips on average than members using a significance level of 5%.
Non-members seem to take trips about 2 minutes longer than members.

```{r}
t.test(avg ~ mem, data = df_main, var.equal = TRUE)
```

We can also conclude that members take more total trips than non-members
using a significance level of 5%.

```{r}
t.test(n_tot ~ mem, data = df_main, var.equal = TRUE)
```

Looking at days when it rains, we can conclude that there is a
significant difference in average trip duration on rainy days compared
to non-rainy days with a 95% confidence level. Though the difference is
by less than 1 minute.

```{r}
t.test(avg ~ rain_ind, data = df_main, var.equal = TRUE)
```

We can also conclude that the mean difference in number of trips in
rainy days and non-rainy days is significant using a 5% significance
level.

```{r}
t.test(n_tot ~ rain_ind, data = df_main, var.equal = TRUE)
```

Through the above t-test, we can expect holidays periods to play a less
important role in our variables of interest and for weather as well as
membership to play a more significant role. As such we can explore
strategies such as dynamic pricing given the weather and membership
incentives.

# Research Question 1: How does membership and holidays affect likelihood of longer trips (average duration exceeds 15 minutes) ?

**Objective of Analysis:** Understand member behavior in terms of rental
duration to tailor membership benefits and pricing strategies.

## Variables Selection

To evaluate how the membership and holidays affect likelihood of longer
trips, we need to create a model that accounts for the membership and
holidays variable and to use the new variable created to know if the
trip is above 15 minutes as the interest variable. The goal would be to
quantify the relationship between these factor and the "longer trips"
variable.

## Model

Here we use a logistic regression model, which is a type of GLM suitable
for binary outcomes, with a binomial distribution and a logit link
function.

```{r model1, echo=FALSE}
glm_membership <- glm(avg_15_ind ~ mem + holiday + wknd_ind, family = binomial, data = df_main)
summary(glm_membership)
```

## Interpretation

-   **Intercept (`-0.13473`)**: This is the log-odds of a trip exceeding
    15 minutes when all predictors (mem, holiday, and wknd_ind) are at
    their reference level (likely non-members, non-holiday, and on
<<<<<<< HEAD
    weekdays).**Odds**: $e^{-0.13473} = 0.87395$.**Probability
    (Intercept)**: $\frac{0.87395}{1 + 0.87395} = 0.4664$ (approximately
    46.64%). This means that when all independent variables are at their
    reference levels, the probability of a trip exceeding 15 minutes is
    approximately 46.64%.
=======
    weekdays).**Odds ratio (Intercept)**:
    $e^{-0.13473} = 0.87395$.**Probability (Intercept)**:
    $\frac{0.87395}{1 + 0.87395} = 0.4664$ (approximately 46.64%). This
    means when all independent variables are at their reference levels,
    the probability of a trip exceeding 15 minutes is approximately
    46.64%.
>>>>>>> ab399c210b84b4094948cf81133cf30ddb803a6f

-   **Membership (`mem1`, `-0.88328`)**: The log-odds of a trip
    exceeding 15 minutes is 0.88328 units lower for members compared to
    non-members, holding other variables constant (Non-holidays and
<<<<<<< HEAD
    weekdays). **Odds ratio**: $e^{-0.88328} = 0.41342$. This means that
    the odds of a trip exceeding 15 minutes for members are on average
    0.41 times the odds for non-members holding all else constant. When
    all variables are at their reference level except membership, we
    expect a probability of seeing a longer trip of
    $\frac{e^(-0.13473-0.88328)}{1 + e^(-0.13473-0.88328)} = 0.265$

-   **Holiday (`holiday1`, `0.60985`)**: The log-odds of a trip
    exceeding 15 minutes is 0.60985 units higher during holidays
    compared to non-holidays, holding other variables constant. **Odds
    ratio**: $e^{0.60985} = 1.84016$. This indicates that the odds of a
    trip exceeding 15 minutes on holidays are on average 1.84 times
    higher than on non-holidays when other covariates are fixed.
=======
    weekdays). **Odds ratio**: $e^{-0.88328} = 0.41342$. **Probability**:
    $\frac{0.41342}{1 + 0.41342} = 0.2925$ (approximately 29.25%). This
    means that the oods of a trip exceeding 15 minutes for members are
    about 0.41 times the odds for non-members, translating to a
    probability of approximately 29.25%.

-   **Holiday (`holiday1`, `0.60985`)**: The log-odds of a trip
    exceeding 15 minutes is 0.60985 units higher during holidays
    compared to non-holidays, holding other variables constant.
    **Odds ratio**: $e^{0.60985} = 1.84016$. **Probability**:
    $\frac{1.84016}{1 + 1.84016} = 0.6479$ (approximately 64.79%). This
    indicates that the odds of a trip exceeding 15 minutes on holidays
    are about 1.84 times higher than on non-holidays, resulting in a
    probability of approximately 64.79%.
>>>>>>> ab399c210b84b4094948cf81133cf30ddb803a6f

-   **Weekend (`wknd_indWeekend`, `0.55915`)**:The log-odds of a trip
    exceeding 15 minutes is 0.55915 units higher during weekends
    compared to weekdays, holding other variables constant (during
<<<<<<< HEAD
    week-end on non-holidays times). **Odds ratio**:
    $e^{0.55915} = 1.74919$. This means that the odds of a trip
    exceeding 15 minutes on weekends are on average 1.75 times higher
    than on weekdays when other covariates are constant.
=======
    week-end on non-holidays times). **Odds ratio**: $e^{0.55915} = 1.74919$.
    **Probability**: $\frac{1.74919}{1 + 1.74919} = 0.6363$
    (approximately 63.63%). This means that the odds of a trip exceeding
    15 minutes on weekends are about 1.75 times higher than on weekdays,
    translating to a probability of approximately 63.63%.
>>>>>>> ab399c210b84b4094948cf81133cf30ddb803a6f

-   **Statistical Significance**: In this model, all the p-values are
    under the 5% level of significance, indicating that the relationship
    between these variables and the likelihood of taking a longer trip
    are statistically significant.

-   **Model Fit**: The AIC of the model is 12869, which can be used for
    model comparison purposes. The lower the AIC, the better the model
    fits the data while penalizing for complexity.

## Business implications

-   Since members are less likely to take longer trips, membership
    benefits and pricing could be adjusted to encourage more extended
    use, or to better cater to the frequent, shorter trips that members
    seem to prefer.

-   The increase in longer trips during holidays and weekends indicates
    potential opportunities for targeted marketing and promotions to
    encourage bike usage during these periods.

-   The significant increase in the likelihood of longer trips during
    long weekends, especially on the weekend days, suggests that there
    might be a need for increased bike availability and maintenance
    during these times to accommodate the higher demand for leisurely
    rides.

## Anova test

Furthermore, we also performed an Anova test between a full model and
a reduced model. The reduced model in this case is the model from above 
whereas the full model also includes the covariate season. The goal here is to
determine if the addition of season significantly improves the
model's fit. The test statistic is thus the difference in -2 times the
log likelihood (-2LL) of the reduced as well as -2LL of the complete
model with the null hypothesis being that the likelihood is the same and
the alternate being that the likelihood is different. In essence, we
want to determine if the addition of season improves the likelihood
of observing the data.

```{r}
glm_membership_full <- glm(avg_15_ind ~ mem + holiday + wknd_ind + season, family = binomial, data = df_main)
anova(glm_membership,glm_membership_full,test="LRT")
```

-   Model 2 has 9994 degrees of freedom. Model 1 has 9996 degrees of
    freedom, which is two more than Model 2, indicating that two
    additional parameter (here, season(3 levels)) was included in Model 2.
-   Model 1 has a residual deviance of 12861 Model 2 has a residual
    deviance of 12574, which is higher, suggesting a better fit to the
    data than Model 1.
-   The deviance reduction from Model 1 to Model 2 is 286.68 This is
    the difference in the Resid. Dev between the two models,
    representing the improvement in fit due to including the season
    variable.
-   The p-value associated with this deviance reduction is less than
    2.2e-16, which is extremely small and indicates that the improvement
    in the model fit by including season is statistically significant.

The addition of the season variable to the model significantly improves the
model's ability to predict whether a trip will last longer than 15
minutes. The very low p-value associated with the deviance reduction
upon adding season to the model confirms that membership status has a
statistically significant effect on the likelihood of a trip exceeding
15 minutes, beyond what is explained by holiday and weekend indicators
alone. Thus, seasonality will be explored in the next model.

# Research Question 2: How do the seasonal variation in weather affect the total number of trip ?

**Objective of Analysis:** Evaluate how seasonal weather patterns
influence rental numbers to inform seasonal staffing and maintenance
schedules.

## Variables Selection

To evaluate how seasonal weather patterns influence rental numbers, we
need to create a model that accounts for the various factors that can
vary with seasons, such as temperature, rainfall and specific time of
year (season). The goal would be to quantify the relationship between
these factors and the number of rentals, which can then inform decisions
on staffing and maintenance schedules.

## Model

Rental numbers are count data, so a Poisson or negative binomial GLM
would both be suitable. Here, we first fit a Poisson regression model
and check for over-dispersion. Since the value is significantly greater
than 1, we then fit a Negative Binomial model that will be more
appropriate for this particular case.

```{r model_2, echo=FALSE}
# Poisson model
glm_seasonal_weather <- glm(n_tot ~ temp + rain + season, family = poisson, data = df_main)
dispersiontest(glm_seasonal_weather)

# Negative Binomial model
glm_seasonal_weather_2 <- glm.nb(n_tot ~ temp + rain + season, data = df_main)
summary(glm_seasonal_weather_2)
```

## Interpretation

-   **Theta**: The estimate for theta ($\theta = 0.9902$) is close to 1.
    This indicates that the variance is slightly greater than the mean,
    which is consistent with some overdispersion in the data. This
    overdispersion justifies the use of the Negative Binomial model over
    the Poisson model.

-   **Model Fit**: The AIC of the model is 79919, which can be used for
    model comparison purposes. The lower the AIC, the better the model
    fits the data while penalizing for complexity.

-   **Intercept**: The intercept ($\beta = 2.581791$) represents the log
    count of the total number of trips when all other variables are
    zero. This is on a day in fall where the temperature is zero and
    there is no rain.

-   **Temperature (`temp`)**: The incidence rate ratio (IRR) for
    temperature is $e^{0.031157}$, which is approximately 1.032. This
    means that for each one-degree Celsius increase in temperature, the
    expected number of total trips increases by a factor of 1.032, or
    3.2%.

-   **Rainfall (`rain`)**: The IRR for rainfall is $e^{-0.015916}$,
    which is approximately 0.984. This indicates that for each
    additional millimeter of rainfall, the expected number of trips
    decreases by a factor of 0.984, or 1.6%. So, if rainfall increases
    by 1 mm, the model predicts a 1.6% decrease in the number of trips
    holding all else constant.

-   **Season (`seasonSpring`, `seasonSummer`)**:

    -   For `seasonSpring`: The IRR is $e^{-0.265988}$, approximately
        0.767. This suggests that, all else being equal, the expected
        number of trips in spring is 76.7% of the number in the baseline
        season (`seasonFall`), which is a 23.3% decrease.

    -   For `seasonSummer`: The IRR is $e^{-0.113913}$, approximately
        0.892. This means that in summer, the expected number of trips
        is 89.2% of the number in the baseline season (`seasonFall`), a
        10.8% decrease.

## Business implications

-   The bike-sharing service is likely to see increased demand on
    warmer, drier days. This can guide the allocation of bikes across
    stations and the scheduling of staff for redistribution and customer
    service.

-   During rainy days, demand is expected to drop, which could be a good
    time for scheduling maintenance work.

-   The observed reduction in trips during the spring and summer, when
    compared to the baseline season, indicates that there might be other
    influencing factors to consider. This observation is made under the
    assumption that temperature remains constant, which is unlikely to
    be accurate. In reality, trip frequency is probably not lower in the
    summer than in the fall. Therefore, it might be necessary to explore
    additional variables or implement targeted marketing strategies to
    enhance ridership in these seasons.

    ##  Anova test

```{r}
# Poisson model
glm_seasonal_weather_full <- glm.nb(n_tot ~ mem + holiday + temp + rain + season, data = df_main)
anova(glm_seasonal_weather,glm_seasonal_weather_full,test="LRT")
```

-   Model 1: Predicts n_tot using temp, rain, and season as predictors.

-   Model 2: Predicts n_tot using mem, holiday, temp, rain, and season.

-   Degrees of Freedom (Df): Model 1 has 9995 residual degrees of
    freedom, and Model 2 has 9993, which suggests that two parameters
    were added in Model 2 (mem and holiday).

-   Residual Deviance: It is a measure of unexplained variance by the
    model. Model 1 has a residual deviance of 204541, and Model 2 has a
    significantly lower residual deviance of 125397, indicating that
    Model 2 fits the data much better.

-   Deviance Difference: The difference in deviance between the two
    models is 79144, which is highly significant (p \< 2.2e-16),
    indicating that the predictors added in Model 2 (mem and holiday)
    significantly improve the model.

In conclusion, the analysis strongly suggests that including mem and
holiday improves the model's ability to predict n_tot. Given that
information, some additional research on the member variable impact on
the number of trips will be made. Additionally, the presence of
overdispersion justifies the use of the Negative Binomial model over the
Poisson model for this data.

# Research Question 3: What variables impacts the porportion of trips in the morning versus in the evening and in what way ?

**Objective of Analysis:** The goal of this analysis is to understand
what variables influence the repartition of the trips throughout the
day. Knowing this would help to better forecast the demand for bikes
across the bixi system.

Before starting the analysis, it is important to know that our datasets
has 57.76% of its trips in the afternoon.

```{r}
sum(df_main$n_PM)/sum(df_main$n_tot)
```

## Variables Selection

Some variables that would be interesting to investigate are the
following:

`mem` : Membership indicator

`wknd_ind` : Indicator of weekend

`season` : Categorical variable with autumn, summer and fall

`temp` : Temperature in degrees celcius

`rain` : Precipitation in mm

`North_South` : Indicator of cardinality compared to parc lafontaine

`West_East`: Indicator of cardinality compared to parc lafontaine

`Metro_ind` : Indicator of metro station nearby

**Correlation:**

The correlation between the variables selected above was already tested
via multicollinearity tests like VIF in previous analysis and did not
show source of concern.

**Interest variable:**

Our interest variable for this question is the proportion of trips in
the afternoon compared to the total number of trips. Since `n_PM` is a
count we expect a poisson like distribution. To obtain a rate we will
use the variable `n_tot` as an offset.

```{r interest variable}
hist(df_main$n_PM, breaks = 30)
```

```{r dispersion}
mean(df_main$n_PM)
var(df_main$n_PM)
```

We observe some big disparities between the mean and variance of the
variable `n_PM` which could lead to some overdispersion in our model.
Some formal test will be explored in the model part.

## Model

```{r model_3, echo=FALSE}
mod.poi <- glm(n_PM ~ mem + wknd_ind +season + temp + rain + North_South  + West_East + Metro_ind + offset(log(n_tot)) , data = df_main, family = poisson(link = "log"))

summary(mod.poi)

```

```{r}
# Deviance based
mod.poi$deviance/mod.poi$df.residual

# Based on Pearon X2 statistic
sum(residuals(mod.poi, type = "pearson")^2)/mod.poi$df.residual
```

Once the covariates and offset are taken into consideration, we seem to
be more in a case of underdispersion since the dispersion (k) is smaller
than 1. For this reason, we will explore a quasipoisson distribution in
order to increase flexibility and allow the mean to be different from
the variance.

```{r}
mod.quasi <- glm(n_PM ~ mem + wknd_ind +season + temp + rain + North_South  + West_East + Metro_ind + offset(log(n_tot)) , data = df_main, family = quasipoisson)
summary(mod.quasi)
```

```{r}
Anova(mod.quasi, type = 3)
```

## Interpretation

**Overall Model** We observe that the dispersion parameter of the model
is estimated to be 0.68 which means that the means given the covariates
is bigger than its variance. According to the Anova command, all
variables included in the model are significant, once adjusted for all
the covariates, except for `metro_ind`

**Intercept** : -0.605 can be interpreted as the log average rate of
trip in the afternoon when all covariates have values of zero.

In other words, this is the average rate of trip in the afternoon for
non-member, during the weekday, in fall, when temperature is zero
degrees celcius, no precipitation, with rental location north east of
parc lafontaine and at an acces point that as no metro station nearby
which is exp(-0.605) = 54.6%.

**Membership :** 0.039 is a positive coefficient hence membership
increase the rate of trip in the afternoon, on average, when all else is
being held constant by a factor of exp(0.039) = 1.039 which means an
increase of about 4% .

**Weekend:** 0.022 is a positive coefficient hence weekend increase the
rate of trip in the afternoon, on average, when all else is being held
constant by a factor of exp(0.022) = 1.022 which means an increase of
about 2% .

**Season:** Spring's coefficient is 0.181 and summer's is 0.005 meaning
that they both have an increase in rate of trip in the afternoon on
average compared to fall when all else is being held constant. This
increase is of a factor of exp(0.181) = 1.198 and exp(0.005) = 1.005
respectively for both season.

**Temperature:** -0.002 which means that a one degree celcius increase
in temperature results in a decrease in rate of afternoon trips on
average when all else is being held constant. This decrease is of a
factor of exp(-0.002) = 0.998.

**Precipitation :**- rain coefficient is also -0.002 hence its
interpretation is the same as for temperature except that the decrease
is for each additional milliliters of rain.

**Cardinality North-South:** 0.041 is a positive coefficient hence
departure from bixi station south of parc lafontaine have a higher rate
of trip in the afternoon, on average, when all else is being held
constant by a factor of exp(0.041) = 1.041 which means an increase of
about 4% compared to northern departure.

**Cardinality West-East:** -0.017 is a positive coefficient hence
departure from bixi station west of parc lafontaine have a lower rate of
trip in the afternoon, on average, when all else is being held constant
by a factor of exp(-0.017) = 0.983 which means a decrease of about 2%
compared to eastern departure.

**Metro station nearby:** coefficient is 0.009 which although not being
significatively differrent from zero can be interpreted as having a
metro station nearby increase the rate of trip in the afternoon, on
average, when all else is being held constant by a factor of exp(0.009)
= 1.009 .

## Business Implications:

The main takeaways from this model are:

-   Members have a higher rate of trips in the afternoon. Knowing that
    members account for most of the trips, it can explain why there is
    more trip in the afternoon in general.

```{r, echo=FALSE}
df_main %>%    group_by(mem) %>%    summarize(n_tot = sum(n_tot))
```

-   There will be an increase demand on the system in the afternoon
    during the weekend and an increase demand on the system in the
    morning during the weekdays. This could reflect the usage of people
    using bixi to commute to work.

-   There is a strong increase in rate of trips in the afternoon during
    the season of spring, this could be seen as an eagerness for bike
    after winter since afternoon trips seems to be more associated with
    leisure than commuting. Another hypothesis would be that during
    spring the mornings are too cold to bike most often.

-   Following the above hypothesis, as temperature increase, there seems
    to be an increase of rate of trip in the morning. Keep in mind that
    this relation is only true for a given season.

-   Finally concerning the general flow of trips, there seems to be a
    higher rate of departure from stations North West to Parc Lafontaine
    in the morning than in the afternoon. Similarly, we have the inverse
    relation for station in the South East. This means that from an
    operational standpoint, there might be some displacement of bikes
    required from stations to stations depending on the moment of the
    day to keep a balanced fleet of bikes all over the system.

# Research Question 4: Are there significant differences in bike trips counts between weekdays and weekends?

**Objective of Analysis:** The objective of the analysis is to quantify
and assess the differences in ridership during weekend and weekdays. The
analysis would generate insights into the patterns of ridership that
could be useful for operational planning, resource allocation, or
service improvements.

## Variables Selection

We can also see from the histogram that the response variable is skewed
to the right. It's clear that linear regression would not be appropriate
in this context. We can fit the Poisson regression model using the glm
function, specifying that the distribution is poisson. We'll start by
fitting the model which includes all explanatory variable:

```{r}
hist(df_main$n_tot, breaks = 20, main = "Histogram of n_tot", xlab = "n_tot")

```

## Model

```{r model_4, echo=FALSE}

# Create a new variable to denote weekdays and weekends
df_main <- df_main %>%
  mutate(weekday_weekend = ifelse(wday %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))

# Fit separate GLMs for trip counts by weekday/weekend
total_model <- glm(n_tot ~ weekday_weekend, data = df_main, family = poisson)

# View summaries of the models
summary(total_model)

```

## Interpretation

**Intercept** :

The intercept (3.004083) represents the expected log count of the total
number of trips on weekdays.In other words, the expected number of trips
on weekdays is exp(3.004083)=20.

**Weekend** :

weekend = -0.040604: number of trips is higher on weekdays than weekend.
More specifically, the mean number of trips on weekend is exp(-0.040604)
= 0.96 times those on weekdays. In other words, the mean number of trips
on weekend is 4% lower than that in weekdays.The p-value is 3.37e-16
which is less that any acceptable value of alpha which represent that
the difference in trip number during weekend and weekdays is
statistically different.

## Business Implications:

**Resource Allocation**: Given the higher demand during weekdays,
businesses or services dependent on these trips may need to allocate
more resources, such as bikes, maintenance, or staff, during these
periods.

**Pricing Strategies**: If the service is fee-based, adjusting pricing
strategies to account for the difference in demand between weekdays and
weekends could be considered. For instance, offering promotions or
discounts during lower demand days (weekends) to attract more users.

**Marketing Efforts**: Tailoring marketing campaigns or efforts to
encourage more weekend usage could be explored. Promoting special
events, family packages, or leisure-oriented offers during weekends
might help in increasing weekend ridership.

**Operational Optimization**: During weekends, operational adjustments
could be made to enhance the user experience. For instance, ensuring
bike availability, adjusting operating hours, or implementing
user-friendly initiatives to attract more weekend users.

**Service Enhancements**: Understanding the differences in usage
patterns can guide service improvements. Addressing any barriers that
might discourage weekend ridership, such as safety concerns, parking
availability, or service accessibility, could be a focus for
enhancements.

# Limitations and shortcomings

**Autocorrelation of data**: The observations in the dataset are not
independent, as seen in previous analyses. This autocorrelation can
impact the validity of statistical tests and models, potentially leading
to biased or inaccurate conclusions.

**External Factors and Generalizability**: The analysis primarily
focuses on internal variables within the dataset, overlooking potential
external influences such as changes in city infrastructure or broader
economic conditions. This limits the generalizability of the results to
broader contexts.

**Temporal Dynamics and Long-Term Trends**: The study's insights are
confined to the timeframe for year 2021, potentially missing long-term
shifts in user behavior or external factors. The temporal dynamics of
the bike-sharing service may evolve beyond the study period.

**Model Specificity**: The use of Generalized Linear Models (GLMs) is
tailored for the specific dataset and research questions at hand. While
GLMs are versatile, they might not capture all the nuances or
complexities of the data, such as non-linear relationships or
interactions between variables that a different modeling approach could
reveal.

**Sensitivity to Parameter Choices**: GLMs involve decisions about which
link function to use and how to structure the model. Different choices
can lead to different interpretations, and the report does not discuss
the sensitivity of the results to these choices.

**Assumptions of the Modeling Approach**: Every statistical model,
including GLMs, comes with underlying assumptions (e.g., about the
distribution of errors). If these assumptions are violated, it can lead
to biased results. The report does not explicitly discuss how these
assumptions were tested or met.

# Conclusion

In conclusion, the examination of Bixi's operational data utilizing
Generalized Linear Models (GLM) has unveiled pivotal insights that can
strategically reshape the bike-sharing service. The analysis has
pinpointed the profound impact of weather conditions, seasonal
variations, and membership dynamics on ridership behavior. This newfound
understanding positions Bixi to implement targeted strategies,
optimizing resource allocation, and addressing specific user preferences
to enhance the overall service experience.

Strategically, Bixi is poised to benefit from dynamic resource
allocation informed by weather patterns, ensuring optimal bike
distribution and efficient staff scheduling. The identification of
seasonal ridership nuances prompts tailored marketing initiatives,
providing an opportunity to counteract dips in colder weather ridership.
Moreover, the strategic refinement of membership structures aligns with
user preferences, enhancing engagement and satisfaction. Capitalizing on
the surge in longer trips during holidays and extended weekends through
targeted campaigns further positions Bixi to maximize user engagement
and solidify its position in the competitive urban mobility sector.

# Contribution

Charles Julien : Business question 3, correction and integration of all parts

Gabriel Jobert : Business question 1 and 2

Chike Odenigbo : t-test, anova, added to feature engineering, part of
conclusion

Atul Sharma : Introduction and Research question 4
